---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(arrow)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sparrow

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/paleolimbot/sparrow/branch/master/graph/badge.svg)](https://codecov.io/gh/paleolimbot/sparrow?branch=master)
[![R-CMD-check](https://github.com/paleolimbot/sparrow/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/sparrow/actions)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of sparrow is to wrap the [Arrow Data C API](https://arrow.apache.org/docs/format/CDataInterface.html) and [Arrow Stream C API](https://arrow.apache.org/docs/format/CStreamInterface.html) to provide lightweight Arrow support for R packages to consume and produce streams of data in Arrow format.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("paleolimbot/sparrow")
```

## Creating arrays

You can create an Arrow array using `as_sparrow_array()`. For many types (e.g., integers and doubles), this is done without any copying of memory: sparrow just arranges the existing R vector memory and protects it for the lifetime of the underlying `struct ArrowArray`.

```{r example}
library(sparrow)
(array <- as_sparrow_array(1:5))
```

For `Array`s and `RecordBatch`es from the [arrow package](https://arrow.apache.org/docs/r/), this is almost always a zero-copy operation and is instantaneous even for very large Arrays.

```{r}
library(arrow)
(array2 <- as_sparrow_array(Array$create(1:5)))
```

## Exporting arrays

To convert an array object to some other type, use `from_sparrow_array()`:

```{r}
str(from_sparrow_array(array))
```

The sparrow package has built-in defaults for converting arrays to R objects; you can also specify your own using the `ptype` argument:

```{r}
str(from_sparrow_array(array, ptype = double()))
from_sparrow_array(array, ptype = arrow::Array)
```

## Streams

The Arrow C API also specifies an experimental stream interface. In addition to handling streams created elsewhere, you can create streams based on a `sparrow_array()`:

```{r}
stream1 <- as_sparrow_array_stream(as_sparrow_array(1:3))
sparrow_array_stream_get_next(stream1)
sparrow_array_stream_get_next(stream1)
```

...or based on a function that returns one or more `sparrow_array()`s:

```{r}
counter <- -1
rows_per_chunk <- 5
csv_file <- readr::readr_example("mtcars.csv")
schema <- as_sparrow_array(
  readr::read_csv(
    csv_file,
    n_max = 0,
    col_types = readr::cols(.default = readr::col_double())
  )
)$schema

stream2 <- sparrow_array_stream_function(schema, function() {
  counter <<- counter + 1L
  result <- readr::read_csv(
    csv_file,
    skip = 1 + (counter * rows_per_chunk),
    n_max = rows_per_chunk,
    col_names = c(
      "mpg", "cyl", "disp", "hp", "drat",
      "wt", "qsec", "vs", "am", "gear", "carb"
    ),
    col_types = readr::cols(.default = readr::col_double())
  )

  if (nrow(result) > 0) result else NULL
})
```

You can pass these to Arrow as a `RecordBatchReader` using `sparrow_array_stream_to_arrow()`:

```{r}
reader <- sparrow_array_stream_to_arrow(stream2)
as.data.frame(reader$read_table())
```

Currently attemping to export an arrow `RecordBatchReader()` segfaults for an unknown reason, but in theory one could also go the other direction.

## C data access

The C data interface is ABI stable (and a version of the stream interface will be ABI stable in the future) so you can access the underlying pointers in compiled code from any R package (or inline C or C++ code). A `sparrow_schema()` is an external pointer to a `struct ArrowSchema`, a `sparrow_array_data()` is an external pointer to a `struct ArrowArray`, and a `sparrow_array()` is a `list()` of a `sparrow_schema()` and a `sparrow_array_data()`.

```{r, include=FALSE}
Sys.setenv(
  PKG_CPPFLAGS = paste0("-I", system.file("include", package = "sparrow"))
)
```

```{c, results='hide'}
#include <R.h>
#include <Rinternals.h>
#include "sparrow.h"

SEXP extract_null_count(SEXP array_data_xptr) {
  struct ArrowArray* array_data = (struct ArrowArray*) R_ExternalPtrAddr(array_data_xptr);
  return Rf_ScalarInteger(array_data->null_count);
}
```

```{r}
.Call("extract_null_count", as_sparrow_array(c(NA, NA, 1:5))$array_data)
```

The lifecycles of objects pointed to by the external pointers are managed by R's garbage collector: any object that gets garbage collected has its `release()` callback called (if it isn't `NULL`) and the underlying memory for the `struct Arrow...` freed. You can call the `release()` callback yourself from compiled code but you probably don't want to unless you're explicitly limiting access to an object.
