---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# arrowvctrs

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/paleolimbot/arrowvctrs/branch/master/graph/badge.svg)](https://codecov.io/gh/paleolimbot/arrowvctrs?branch=master)
[![R-CMD-check](https://github.com/paleolimbot/arrowvctrs/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/arrowvctrs/actions)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of arrowvctrs is to wrap the [Arrow Data C API](https://arrow.apache.org/docs/format/CDataInterface.html) and [Arrow Stream C API](https://arrow.apache.org/docs/format/CStreamInterface.html) to provide lightweight Arrow support for R packages to consume and produce streams of data in Arrow format. Right now it's just a fun way for me to learn about Arrow!

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("paleolimbot/arrowvctrs")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(arrowvctrs)
(vctr <- as_carrow_array(ggplot2::mpg))
```

```{r}
tibble::as_tibble(from_carrow_array(vctr))
```
